\bf A new class of measurement error models for CRISPR genome engineering and single-cell RNA sequencing

### 2021-12-29 text/jasa/jasa.tex
reformat text for journal

% set document class
\documentclass[11pt]{article}

% load packages
\usepackage{setspace}
\usepackage[margin=1in]{geometry}
\usepackage{amsfonts}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{array}
\usepackage{titlesec} 
\usepackage{xcolor}
\usepackage{float}
\usepackage[toc,page]{appendix}
\usepackage{algpseudocode}
\usepackage{afterpage}
\usepackage[
backend=biber,
style=numeric-comp,
maxcitenames=1,
maxbibnames=3,
backref=true,
doi=false,
isbn=false,
url=false,
eprint=false
]{biblatex}

% set options
\doublespacing
\addbibresource{/Users/timbarry/optionFiles/glmeiv.bib}
\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\newcolumntype{C}[1]{>{\PreserveBackslash\centering}p{#1}}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\makeatletter
\newcommand{\multiline}[1]{%
	\begin{tabularx}{\dimexpr\linewidth-\ALG@thistlm}[t]{@{}X@{}}
		#1
	\end{tabularx}
}
\makeatother
\setcounter{tocdepth}{2}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\allowdisplaybreaks
\usepackage[caption = false]{subfig}
\usepackage{/Users/timbarry/optionFiles/mymacros}
\DeclareMathOperator*{\argmin}{arg\,min}
\titleformat{\subsection}[runin]
{\normalfont\large\bfseries}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[runin]
{\normalfont\normalsize\bfseries}{\thesubsubsection}{1em}{}
\newcommand\blankpage{%
	\null
	\thispagestyle{empty}%
	\addtocounter{page}{-1}%
	\newpage}

\begin{document}
\noindent
\begin{center} 
\textbf{A class of exponential family measurement error models for CRISPR genome engineering and single-cell sequencing} \\
Timothy Barry, Eugene Katsevich, Kathryn Roeder
\end{center}

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. Despite their promise, single-cell CRISPR screens present substantial statistical challenges. We demonstrate through theoretical and real data analyses that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. To overcome these limitations, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis. GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables. We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, demonstrating improved performance in challenging problem settings.
\end{abstract}

\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

@@

The genomics literature has produced several applied methods for single-cell CRISPR screen analysis. In a prior work we developed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we account for the fact that the perturbation is measured with noise, an issue that SCEPTRE circumvents via thresholding. Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.

@@

Let $n \in \N$ be the number of cells assayed in the experiment. Consider a single perturbation and a single gene. For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced; let $g_i \in \N$ be the number of gRNA transcripts sequenced; let $d^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (i.e., the library size or sequencing depth); and finally, let $z_i \in \R^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively. The thresholding method is defined as follows:
\begin{itemize}
\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$ \hat{p}_i = 0 \textrm{ if } g_i < c; \quad \hat{p}_i = 1 \textrm{ if } g_i \geq c.$$
%$$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i < c, \\ \hat{p}_i = 1 \textrm{ if } g_i \geq c. \end{cases}$$
\item[2.] Assume that $m_i$ is related to $\hat{p}_i, d^m_i,$ and $z_i$ through the following GLM:
\begin{equation}\label{thresh_glm}
m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i); \quad
\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right),
\end{equation}
where (i) $NB_{\theta^m}(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.

@@

Next, we studied the thresholding method from a theoretical perspective, recovering in precise mathematical terms phenomena revealed in the empirical analysis (plus several others). We worked in a simplified Gaussian setting. Suppose we observe gRNA expression and gene expression data $(g_1, m_1), \dots, (g_n, m_n)$ on $n$ cells from the following linear model:
\begin{equation}\label{theoretical_model}
m_i = \beta^m_0 + \beta^m_1 p_i + \ep_i; \quad
g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i; \quad
p_i \sim \textrm{Bern}(\pi); \quad
\ep_i, \tau_i \sim N(0,1),
\end{equation}
where $p_i, \tau_i,$ and $\ep_i$ are independent. For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ is the OLS solution, i.e. $\hat{\beta}^m_1 = \left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})(m_i - \overline{m})\right]\left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2\right]^{-1}.$ We derive the almost sure limit of $\hat{\beta}^m_1$: 
\begin{proposition}\label{prop:convergence}
	The almost sure limit (as $n \to \infty$) of $\hat{\beta}^m_1$ is
	\begin{equation}\label{thresh_est_intercepts}
	\hat{\beta}^m_1 \xrightarrow{a.s.} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right) \equiv \beta^m_1 \gamma(\beta^g_1, \pi, c, \beta^g_0), 
	\end{equation} where 
	$$
	\E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi; \quad \omega = \Phi\left(\beta_1^g + \beta_0^g -c \right); \quad \zeta = \Phi\left( \beta^g_0 - c \right).$$
\end{proposition}
The function $\gamma: \R^4 \to \R$ does not depend on the gene expression parameters $\beta^m_1$ or $\beta^m_0$. The asymptotic relative bias $b: \R^4 \to \R$ of $\hat{\beta}^m_1$ is given by
$$b(\beta^g_1, \pi, c, \beta^g_0)  \equiv \frac{1}{\beta^m_1} \left(\beta^m_1 - \lim_\textrm{a.s.} \hat{\beta}^m_1 \right) = 1 - \gamma(\beta^g_1, \pi, c, \beta^g_0).$$
Having derived an exact expression for  the asymptotic relative bias of  $\hat{\beta}^m_1$, we can prove several results about this quantity. For the sake of simplicity, we consider a slightly specialized setting in which we set $\pi$ to to a specific value (see Appendix \ref{sec:appendix_theory} for details; proofs and detailed statements likewise deferred to Appendix \ref{sec:appendix_theory}).

First, the thresholding estimator strictly \textit{underestimates} (in absolute value) the true value of $\beta^m_1$ over all choices of the threshold and over all possible values of the model parameters. This phenomenon, called attenuation bias, is a common attribute of estimators that ignore measurement in errors-in-variables models \cite{Stefanski2000a}. Second, the magnitude of the bias decreases monotonically in $\beta^g_1$, comporting with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated. Third, the Bayes-optimal decision boundary $c_\textrm{bayes} \in \R$ (i.e., the most accurate decision boundary for classifying cells as perturbed or unperturbed) is a critical value of the bias function. Finally, and most subtly, there is no universally applicable rule for selecting a threshold that produces minimal bias: when $\beta^g_1$ is small, setting the threshold to an arbitrarily large number yields smaller bias than setting the threshold to the Bayes decision boundary; when $\beta^g_1$ is large, the reverse is true.

Next, we studied the variance of the thresholding estimator, considering a slightly simpler model for this purpose. Suppose the intercepts in (\ref{theoretical_model}) are fixed at $0$ (i.e., $\beta^m_0$ = $\beta^g_0$ = 0). For notational simplicity write $\beta_m = \beta^m_1$ and $\beta_g = \beta^g_1.$ The thresholding estimator $\hat{\beta}_m$ is the no-intercept OLS solution $\hat{\beta}_m = \left[\sum_{i=1}^n \hat{p}_i m_i \right]\left[\sum_{i=1}^n \hat{p}_i^2 \right]^{-1}.$ We have the following limiting distributional result.
\begin{proposition}\label{prop:bv_decomp}
	The limiting distribution of $\hat{\beta}_m$ is
	$$\sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m \omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) }{\left(\E[\hat{p}_i]\right)^2} \right),$$ where $$l = \beta_m \omega \pi/[\zeta(1-\pi) + \omega \pi]; \quad
	\E[\hat{p}_i] = \pi \omega + (1-\pi) \zeta; \quad
	\omega = \Phi(\beta_g - c); \quad
	\zeta = \Phi(-c).$$
\end{proposition}
This proposition yields an asymptotically exact bias-variance decomposition for $\hat{\beta}_m$ (Figure \ref{thresholding_theoretical}): as the threshold tends to infinity, the bias decreases and the variance increases, consistent with empirical observations that we made in Section \ref{sec:thresholding_empirical}. Overall, our empirical and theoretical analyses reveal that thresholded regression poses several difficulties: thresholding incurs strict attenuation bias (that in some settings is large), and selecting a good threshold is challenging. These considerations motivate our core research question: \textit{does modeling the gRNA counts directly, thereby circumventing thresholding, facilitate estimation and inference in single-cell CRISPR screen analysis}?

@@

\subsection{Model}\label{sec:model}
Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})). We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities. To this end, let the latent variable $p_i \in \{0,1\}$ indicate whether cell $i \in \{1, \dots, n\}$ was perturbed. We model the gene expression counts according to
\begin{equation}\label{glmeiv_model_1}
m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i^m); \quad \log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
\end{equation}
where $\theta^m > 0$ is a known negative binomial size parameter, and $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in \R^{d - 2}$ are unknown constants. The model (\ref{glmeiv_model_1}) is identical to the thresholding model (\ref{thresh_glm}), but the imputed perturbation indicator $\hat{p}_i$ is replaced by the latent perturbation indicator $p_i$. Next, let $d^g_i \in \N$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size). The model for the gRNA counts is 
\begin{equation}\label{glmeiv_model_2}
g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right); \quad \log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
\end{equation}
where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \R, \beta^g_1 \in \R, \gamma_g \in \R^{d - 2}$ are unknown constants. We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \cite{Datlinger2017,Hill2018}. Finally, we model the marginal perturbation probability as $
p_i \sim \textrm{Bern}(\pi),$ where $\pi \in (0,1/2]$. 
 Together, (\ref{glmeiv_model_1}), (\ref{glmeiv_model_2}), and the marginal distribution of $p_i$ define the negative binomial GLM-EIV model. The terms $(\beta^m_0 + \beta^m_1p_i + \gamma^T_m z_i)$ and $(\beta^g_0 + \beta^g_1 p_i + \gamma^T_g z_i)$ can be interpreted as relative gene and gRNA expressions, similar to the analogous term in the thresholding model. Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors. See Appendix \ref{sec:zero_inf_model} for a parallel methodology that we developed in which we model the gRNA distribution as (in a certain sense) zero-inflated counts that may be applicable in some settings.

\subsubsection*{General model}
To provide greater modeling flexibility, we generalize the GLM-EIV model to arbitrary exponential family response distributions and link functions. To increase notational compactness, let $\tilde{x_i} = [1, p_i, z_i]^T \in \R^d$ be the vector of covariates (including an intercept term) for the $i$th cell. (We use the tilde as a reminder that the vector is partially unobserved.) Let $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \R^d$ and $\beta_g = [\beta^g_0, \beta^g_1, \gamma_g]^T \in \R^d$ be the unknown coefficient vectors corresponding to the gene and gRNA expression models, respectively. Finally, let $o^m_i$ and $o^g_i$ be the (possibly zero) offset terms for the gene and gRNA models; in practice, we typically set $o^m_i$ and $o^g_i$ to the log-transformed library sizes (i.e., $\log(d^m_i)$ and $\log(d^g_i)$, respectively).

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{setspace}
\usepackage{amsfonts}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{array}
\usepackage{titlesec} 
\usepackage{xcolor}
\usepackage{float}
\usepackage[toc,page]{appendix}
\usepackage{algpseudocode}
\usepackage{afterpage}
\usepackage[
backend=biber,
style=numeric-comp,
maxcitenames=1,
maxbibnames=3,
backref=true,
doi=false,
isbn=false,
url=false,
eprint=false
]{biblatex}

% MY COMMANDS
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\addbibresource{/Users/timbarry/optionFiles/glmeiv.bib}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\newcolumntype{C}[1]{>{\PreserveBackslash\centering}p{#1}}
\newcommand{\indep}{\raisebox{0.05em}{\rotatebox[origin=c]{90}{$\models$}}}

\makeatletter
\newcommand{\multiline}[1]{%
	\begin{tabularx}{\dimexpr\linewidth-\ALG@thistlm}[t]{@{}X@{}}
		#1
	\end{tabularx}
}
\makeatother

% JOURNAL SETTINGS
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}
\addtolength{\oddsidemargin}{-.5in}
\addtolength{\evensidemargin}{-1in}
\addtolength{\textwidth}{1in}
\addtolength{\textheight}{1.7in}
\addtolength{\topmargin}{-1in}
\date{\vspace{-5ex}}

\begin{document}
\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}
\if1\blind
{
  \title{\bf A new class of measurement error models for CRISPR genome engineering and single-cell RNA sequencing}
  \author{Timothy Barry$^1$, Eugene Katsevich$^2$, Kathyrn Roeder$^1$ \\
    $^1$Department of Statistics and Data Science, Carnegie Mellon University \\
    $^2$Department of Statistics and Data Science, Wharton School}
  \maketitle
} \fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE \bf A new class of measurement error models for CRISPR genome engineering and single-cell RNA sequencing}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
CRISPR genome engineering and single-cell RNA sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. Despite their promise, single-cell CRISPR screens present substantial statistical challenges. We demonstrate through theoretical and real data analyses that a standard method for estimation and inference in single-cell CRISPR screens --``thresholded regression'' -- exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic and challenging-to-select tuning parameter. To overcome these difficulties, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis. GLM-EIV extends the classical errors-in-variables model to responses and noisy predictors that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables. We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, demonstrating improved performance in challenging problem settings.
\end{abstract}

\noindent
{\it Keywords:}  exponential family, errors-in-variables, GLM, CRISPR screen, differential expression
\spacingset{1.9}
\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

@@

The genomics literature has produced several applied methods for single-cell CRISPR screen analysis. Barry et al.\ developed SCEPTRE \cite{Barry2021}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we account for the fact that the perturbation is measured with noise, an issue that SCEPTRE circumvents via thresholding. Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.

@@

Let $n \in \mathbb{N}$ be the number of cells assayed in the experiment. Consider a single perturbation and a single gene. For cell $i \in \{1, \dots, n\}$, let $m_i \in \mathbb{N}$ be the number of gene transcripts sequenced; let $g_i \in \mathbb{N}$ be the number of gRNA transcripts sequenced; let $d^m_i \in \mathbb{N}$ be the number of gene transcripts sequenced across \textit{all} genes (i.e., the library size or sequencing depth); and finally, let $z_i \in \mathbb{R}^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively. The thresholding method is defined as follows:
\begin{itemize}
	\item[1.] For a given threshold $c \in \mathbb{N}$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$ \hat{p}_i = 0 \textrm{ if } g_i < c; \quad \hat{p}_i = 1 \textrm{ if } g_i \geq c.$$
	%$$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i < c, \\ \hat{p}_i = 1 \textrm{ if } g_i \geq c. \end{cases}$$
	\item[2.] Assume that $m_i$ is related to $\hat{p}_i, d^m_i,$ and $z_i$ through the following GLM:
	\begin{equation}\label{thresh_glm}
	m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i); \quad
	\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right),
	\end{equation}
	where (i) $NB_{\theta^m}(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \mathbb{R}, \beta^m_1 \in \mathbb{R},$ and $\gamma_m \in  \mathbb{R}^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.

@@

Next, we studied the thresholding method from a theoretical perspective, recovering in precise mathematical terms phenomena revealed in the empirical analysis (plus several others). We worked in a simplified Gaussian setting. Suppose we observe gRNA expression and gene expression data $(g_1, m_1), \dots, (g_n, m_n)$ on $n$ cells from the following linear model:
\begin{equation}\label{theoretical_model}
m_i = \beta^m_0 + \beta^m_1 p_i + \epsilon_i; \quad
g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i; \quad
p_i \sim \textrm{Bern}(\pi); \quad
\epsilon_i, \tau_i \sim N(0,1),
\end{equation}
where $p_i, \tau_i,$ and $\epsilon_i$ are independent. For a given threshold $c \in \mathbb{R}$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ is the OLS solution, i.e. $\hat{\beta}^m_1 = \left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})(m_i - \overline{m})\right]\left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2\right]^{-1}.$ We derive the almost sure limit of $\hat{\beta}^m_1$: 
\begin{proposition}\label{prop:convergence}
	The almost sure limit (as $n \to \infty$) of $\hat{\beta}^m_1$ is
	\begin{equation}\label{thresh_est_intercepts}
	\hat{\beta}^m_1 \xrightarrow{a.s.} \beta^m_1 \left(\frac{ \pi( \omega - \mathbb{E}[ \hat{p}_i ])}{ \mathbb{E}[\hat{p}_i] (1 - \mathbb{E}[\hat{p}_i])}\right) \equiv \beta^m_1 \gamma(\beta^g_1, \pi, c, \beta^g_0), 
	\end{equation} where 
	$$
	\mathbb{E}[\hat{p}_i] = \zeta(1-\pi) + \omega\pi; \quad \omega = \Phi\left(\beta_1^g + \beta_0^g -c \right); \quad \zeta = \Phi\left( \beta^g_0 - c \right).$$
\end{proposition}
The function $\gamma: \mathbb{R}^4 \to \mathbb{R}$ does not depend on the gene expression parameters $\beta^m_1$ or $\beta^m_0$. The asymptotic relative bias $b: \mathbb{R}^4 \to \mathbb{R}$ of $\hat{\beta}^m_1$ is given by
$$b(\beta^g_1, \pi, c, \beta^g_0)  \equiv \frac{1}{\beta^m_1} \left(\beta^m_1 - \lim_\textrm{a.s.} \hat{\beta}^m_1 \right) = 1 - \gamma(\beta^g_1, \pi, c, \beta^g_0).$$
Having derived an exact expression for  the asymptotic relative bias of  $\hat{\beta}^m_1$, we can prove several results about this quantity. For the sake of simplicity, we consider a slightly specialized setting in which we set $\pi$ to to a specific value (see Appendix \ref{sec:appendix_theory} for details; proofs and detailed statements likewise deferred to Appendix \ref{sec:appendix_theory}).

First, the thresholding estimator strictly \textit{underestimates} (in absolute value) the true value of $\beta^m_1$ over all choices of the threshold and over all possible values of the model parameters. This phenomenon, called attenuation bias, is a common attribute of estimators that ignore measurement in errors-in-variables models \cite{Stefanski2000a}. Second, the magnitude of the bias decreases monotonically in $\beta^g_1$, comporting with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated. Third, the Bayes-optimal decision boundary $c_\textrm{bayes} \in \mathbb{R}$ (i.e., the most accurate decision boundary for classifying cells as perturbed or unperturbed) is a critical value of the bias function. Finally, and most subtly, there is no universally applicable rule for selecting a threshold that produces minimal bias: when $\beta^g_1$ is small, setting the threshold to an arbitrarily large number yields smaller bias than setting the threshold to the Bayes decision boundary; when $\beta^g_1$ is large, the reverse is true.

Next, we studied the variance of the thresholding estimator, considering a slightly simpler model for this purpose. Suppose the intercepts in (\ref{theoretical_model}) are fixed at $0$ (i.e., $\beta^m_0$ = $\beta^g_0$ = 0). For notational simplicity write $\beta_m = \beta^m_1$ and $\beta_g = \beta^g_1.$ The thresholding estimator $\hat{\beta}_m$ is the no-intercept OLS solution $\hat{\beta}_m = \left[\sum_{i=1}^n \hat{p}_i m_i \right]\left[\sum_{i=1}^n \hat{p}_i^2 \right]^{-1}.$ We have the following limiting distributional result.
\begin{proposition}\label{prop:bv_decomp}
	The limiting distribution of $\hat{\beta}_m$ is
	$$\sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m \omega\pi(\beta_m - 2l) + \mathbb{E}[\hat{p}_i](1 + l^2) }{\left(\mathbb{E}[\hat{p}_i]\right)^2} \right),$$ where $$l = \beta_m \omega \pi/[\zeta(1-\pi) + \omega \pi]; \quad
	\mathbb{E}[\hat{p}_i] = \pi \omega + (1-\pi) \zeta; \quad
	\omega = \Phi(\beta_g - c); \quad
	\zeta = \Phi(-c).$$
\end{proposition}
This proposition yields an asymptotically exact bias-variance decomposition for $\hat{\beta}_m$ (Figure \ref{thresholding_theoretical}): as the threshold tends to infinity, the bias decreases and the variance increases, consistent with empirical observations that we made in Section \ref{sec:thresholding_empirical}. Overall, our empirical and theoretical analyses reveal that thresholded regression poses several difficulties: thresholding incurs strict attenuation bias (that in some settings is large), and selecting a good threshold is challenging. These considerations motivate our core research question: \textit{does modeling the gRNA counts directly, thereby circumventing thresholding, facilitate estimation and inference in single-cell CRISPR screen analysis}?

@@

\subsection{Model}\label{sec:model}
Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})). We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities. To this end, let the latent variable $p_i \in \{0,1\}$ indicate whether cell $i \in \{1, \dots, n\}$ was perturbed. We model the gene expression counts according to
\begin{equation}\label{glmeiv_model_1}
m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i^m); \quad \log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
\end{equation}
where $\theta^m > 0$ is a known negative binomial size parameter, and $\beta^m_0 \in \mathbb{R}, \beta^m_1 \in \mathbb{R},$ and $\gamma_m \in \mathbb{R}^{d - 2}$ are unknown constants. The model (\ref{glmeiv_model_1}) is identical to the thresholding model (\ref{thresh_glm}), but the imputed perturbation indicator $\hat{p}_i$ is replaced by the latent perturbation indicator $p_i$. Next, let $d^g_i \in \mathbb{N}$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size). The model for the gRNA counts is 
\begin{equation}\label{glmeiv_model_2}
g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right); \quad \log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
\end{equation}
where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \mathbb{R}, \beta^g_1 \in \mathbb{R}, \gamma_g \in \mathbb{R}^{d - 2}$ are unknown constants. We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \cite{Datlinger2017,Hill2018}. Finally, we model the marginal perturbation probability as $
p_i \sim \textrm{Bern}(\pi),$ where $\pi \in (0,1/2]$. 
Together, (\ref{glmeiv_model_1}), (\ref{glmeiv_model_2}), and the marginal distribution of $p_i$ define the negative binomial GLM-EIV model. The terms $(\beta^m_0 + \beta^m_1p_i + \gamma^T_m z_i)$ and $(\beta^g_0 + \beta^g_1 p_i + \gamma^T_g z_i)$ can be interpreted as relative gene and gRNA expressions, similar to the analogous term in the thresholding model. Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors. See Appendix \ref{sec:zero_inf_model} for a parallel methodology that we developed in which we model the gRNA distribution as (in a certain sense) zero-inflated counts that may be applicable in some settings.

\subsubsection*{General model}
To provide greater modeling flexibility, we generalize the GLM-EIV model to arbitrary exponential family response distributions and link functions. To increase notational compactness, let $\tilde{x_i} = [1, p_i, z_i]^T \in \mathbb{R}^d$ be the vector of covariates (including an intercept term) for the $i$th cell. (We use the tilde as a reminder that the vector is partially unobserved.) Let $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \mathbb{R}^d$ and $\beta_g = [\beta^g_0, \beta^g_1, \gamma_g]^T \in \mathbb{R}^d$ be the unknown coefficient vectors corresponding to the gene and gRNA expression models, respectively. Finally, let $o^m_i$ and $o^g_i$ be the (possibly zero) offset terms for the gene and gRNA models; in practice, we typically set $o^m_i$ and $o^g_i$ to the log-transformed library sizes (i.e., $\log(d^m_i)$ and $\log(d^g_i)$, respectively).

We use GLMs to model the gene and gRNA expressions. Considering first the gene expression model, let the $i$th linear component $l^m_i$ of the model be $l^m_i = \langle \tilde{x}_i, \beta_m \rangle + o^m_i.$ Next, let the mean $\mu^m_i$ of the $i$th observation be $r_m(\mu^m_i) = l^m_i,$ where $r_m:\mathbb{R} \to \mathbb{R}$ is a strictly increasing, differentiable link function. Let $\psi_m: \mathbb{R} \to \mathbb{R}$ be the differentiable, cumulant-generating function of the selected exponential family distribution. We can express the canonical parameter $\eta^m_i$ in terms of $\psi_m$ and $r_m$ by
$\eta^m_i = \left([\psi'_m]^{-1} \circ r^{-1}_m\right)(l_i^m) := h_m(l_i^m).$ Finally, let $c_m: \mathbb{R} \to \mathbb{R}$ be the carrying density of the selected exponential family distribution. The density $f_m$ of $m_i$ conditional on the the canonical parameter $\eta_i$ is
$f_m(m_i; \eta^m_i) = \exp\left\{m_i \eta^m_i - \psi_m(\eta^m_i) + c_m(m_i) \right\}.$ The function $c_m$ does not appear in the log likelihood of $m_i$; therefore, the only functions relevant to inference are $\psi_m$ and $r_m$.
